---
title: "ESM 244 Lab 2"
author: "Tom Paschos"
date: "1/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Lab 2 Objectives

- Dealing with function conflicts (function name in multiple packages)
- Ordinal logistic regression
- Principle components analysis example + screeplot + biplot
- Shiny example (Marvel characters)

###0. Attach packages: 

```{r load_packages}

library(tidyverse)
library(janitor)
library(MASS)
library(RColorBrewer)
library(factoextra)
library(shinythemes)
library(ggbiplot)
library(ggrepel)

```

Note: MASS package also has a 'select' function that will mask 'select' in dplyr - we don't want that! There are two things we can do: 

(1) State the package explicitly that the function comes from when using it, e.g. dplyr::select

(2) Generally override that the default select is from MASS as follows: 

    select <- dplyr::select
    
  But you'd have to overwrite this each time.
    
```{r, message = FALSE}

select <- dplyr::select

```

###1. OLR for political party affiliation (Open Science Framework: https://osf.io/8n6s4/) by ethnicity, age, and sex

####a. Get data (pol_party.csv) and do some wrangling

```{r, message = FALSE}

# clean_names creates lowercases and underscores

pol_party <- read_csv("pol_party.csv") %>% 
  clean_names() %>%
  select(participant_id:read_carefully)
  
# A little more wrangling

pol_df <- pol_party %>% 
  select(birth_year, gender, ethnicity, political_party) %>% 
  filter(birth_year > 1900, political_party != "NA", political_party <= 7) %>% 
  mutate(age = 2016 - birth_year)

# How do different characteristics influence likelihood that they ID themselves on this political spectrum?
# For political party: 1 = very liberal, 4 = moderate, 7 = very conservative
# When we're doing OLR, we want our Dependent variable we want our class to be ORDERED FACTOR

pol_df$political_party <- factor(pol_df$political_party, ordered = TRUE, levels = c("1","2","3","4","5","6","7"))

# Check class and levels using class function in console to ensure

# Next, update the gender and ethnicity variables to factors (unordered).  These are our categorical predictor variables.

pol_df$ethnicity <- factor(pol_df$ethnicity)
pol_df$gender <- factor(pol_df$gender)

# Then we can run ordinal logistic regression, but we should do some visualization first.

```

- Exploratory visualization of data

```{r, message = FALSE}

# Using tally
# You can see after running these lines that the survey data is heavily weighted towards white respondents, ie only 2 black respondents ID themselves at Level 6, but we want at least n = 15 for sample size

counts_eth <- pol_df %>% 
  group_by(ethnicity, political_party) %>% 
  tally()

counts_g <- pol_df %>% 
  group_by(gender, political_party) %>% 
  tally()

# Exploratory graphs

ggplot(pol_df, aes(x = political_party)) +
  geom_histogram(aes(fill = ethnicity), stat = "count") +
  scale_fill_brewer(palette = "YlGnBu") +
  facet_wrap(~ethnicity, scales = "free")

```

- Ordinal logistic regression using MASS::polr
# WARNING: THESE DATA ARE PRECARIOUS AND N IS TOO SMALL FOR GROUPS

```{r, message = FALSE}
# Ordinal logistic regression

### THIS IS NOT PUBLISHABLE DATA

pol_model <- polr(political_party ~ age + gender + ethnicity, data = pol_df)

# See model summary. Note: can convert this to p-values if you really MUST. 
summary(pol_model)

# Interpretations: logodds coefficient of 0 = No effect; female is the reference sex; since Gender male is positive that means it's an increase in log odds relative to reference level

# logodds are hard to think about, so we can see exponentiated (odds) coefficients

exp(coef(pol_model))

### 

# Once we've decided on a model, we can always makes predictions using it:

predictions <- predict(pol_model, type = "probs")
df <- data.frame(pol_df, predictions)

```

###2. World bank environment PCA (Principle components analysis), using WorldBank enviro indicators

- Data from the World Bank (<https://data.worldbank.org/topic/environment>)

a. Get the data and do some wrangling

```{r, message = FALSE}

wb_10 <- read_csv("wb_10.csv") %>% 
  select(-X1) %>% 
  column_to_rownames('code')


wb_pca <- prcomp(wb_10[3:8], scale = TRUE) # square brackets specifies a subset, oldschool coding style
summary(wb_pca)

ggbiplot(wb_pca) +
  geom_text_repel(label = rownames(wb_10), size = 2, segment.size = 0.2) +
  theme_bw()

#factoextra package (built to help visualize )
#screeplot

fviz_screeplot(wb_pca)

# See contributions of variables to different components

fviz_contrib(wb_pca, choice = "var", axes = 2, top = 6)

# Create a biplot showing points and vectors

fviz_pca_biplot(wb_pca,
                col.var = "cos2",
                label = "var",
                repel = TRUE)

```

More PCA in R examples: 

Clear R code using ggbiplot: 
<https://www.datacamp.com/community/tutorials/pca-analysis-r>

Good examples using factoextra:
<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/>


###3. Shiny example: Marvel characters (create a new Shiny app)


